Random number generation and applications on GPU’s in R(?)

### 2021-10-24 pkg/gpuRandom/inst/documents/paper1_2021_5/paper1.Rnw
update paper1
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/diseasemapping@2920 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\title{Random number generation and applications on GPU’s in R?}
\Plaintitle{}
\Shorttitle{}

@@

In recent years, parallel computing with R \citep{r2021} has become a very important topic and attracted lots of interest from researchers \citep[see][for a review]{eddelbuettel2021parallel}. Although \proglang{R} is one of the most popular statistical software with many advantages, it has drawbacks in memory usage and computation mode aspects \citep{zhao_2016}. To be more specific, (1) \proglang{R} requires all data to be loaded into the main memory (RAM) and thus can handle a very limited size of data; (2) \proglang{R} is a single-threaded program, it can not effectively use all the computing cores of multi-core processors. Parallel computing is the solution to these drawbacks of \proglang{R}, for an overview of current parallel computing approaches with \proglang{R}, see CRAN Task View by \citet{cran2021} at \url{https://cran.r-project.org/web/views/HighPerformanceComputing.html}. One major way of parallel computing with \proglang{R} is using Graphics Processing Units (GPUs). GPUs can perform thousands of tests simultaneously, which makes them powerful at doing massive parallel computing and they are relatively cheap compared to multicore CPU's. Although there has been a few \proglang{R} packages developed that provide some GPU capability in the past years, they come with some restrictions. Packages such as \pkg{gputools}, \pkg{gpumatrix}, \pkg{cudaBayesreg}, \pkg{rpud} (available on github), are now no longer maintained, the popular \pkg{tensorflow} \citep{tensorflow1} package uses GPU via \proglang{Python}, which makes it hard to be incorporated in a new \proglang{R} package that uses GPU. All of these mentioned packages are restricted to \proglang{R} users with NVIDIA GPUs. \pkg{gpuR} \citep{gpur1} is the only \proglang{R} package with a convenient interface between \proglang{R} and GPUs and it is created for any \proglang{R} user with a GPU device. By utilizing the \pkg{ViennaCL} \citep*{rupp2016viennacl} library, it provides a bridge between \proglang{R} and non-proprietary GPUs through an \proglang{OpenCL} (Open Computing Language) backend, which when combined with \pkg{Rcpp} \citep{rcpp1} gives a building block for other \proglang{R} packages.

Random number generator (RNG) is critical in simulation-based statistical inference, there is not yet an \proglang{R} package that is able to generate random numbers on GPU. This is complicate to do on GPU because multiple streams will be needed for parallel generation of random numbers, and the stream states need to be able to be restored for reproducibility of simulation results. We introduce the \pkg{gpuRandom} \proglang{R} package that leverages the \pkg{gpuR} package, and can generate random numbers on GPU by using the \pkg{clRNG} \citep{l2015clrng} library, and can thus accelerate several types of statistical simulation and modelling.

The remaining sections are organized as follows. 
In Section 2, we introduce streams and the use of streams in work-items on GPU device, and describe and check the functions for generating uniform and Normal random numbers.
In Section 3, we apply GPU-generated uniform random numbers in Monte Carlo simulation for Fisher’s exact test, we parallelise this Monte Carlo simulation task and implement it on GPU, then we provide two real data examples to demonstrate the function usage and test its R performance.
Section 4 simulates simultaneously multiple Gaussian random surfaces by applying GPU-generated Normal random numbers and GPU-computed batches of Mat\'ern covariance matrices, then we plot the generated random surfaces in the end. 
Finally, the paper concludes with a short summary and a discussion in Section 5.

@@

\section{Generating random numbers} \label{}
Random numbers are the basis in statistical simulations methods, and are critical in many other fields. \cite{l2012random} summarized the usual two steps to generate a random variable %from a specified distribution 
in computational statistics: (1) generating independent and identically distributed (i.i.d.) uniform random variables on the interval $(0, 1)$, (2) applying transformations to these i.i.d. $U(0, 1)$ random variables to get samples from the desired distribution. \cite{l2012random} and \cite{robert2004random} present several general transformation methods for generating non-uniform random variables, for example, the most frequently used inverse transform method for generating a random variable $X$ from some arbitrary distribution function $F$; The Box-Muller algorithm \citep{box1958note} for Gaussian random variable generation, and so on, which are all built on uniform random variables. Therefore, uniform random number generators (RNGs) are essential in simulating random numbers from all types of probability distributions. In the rest paper, ``RNG'' refers to uniform pseudo random number generator.

\proglang{clRNG} is an \proglang{OpenCL} library for uniform random number generation, it provides four different RNGs: the MRG31k3p, MRG32k3a, LFSR113, and Philox-4×32-10 generators, these RNGs are actually stream objects of different forms. \pkg{gpuRandom} package uses the MRG31k3p RNG from the \proglang{clRNG} library, making it able to generate random numbers on GPUs. In what follows, we will illustrate how to create stream objects and how to use streams to generate (standard) Uniform and (standard) Gaussian random numbers.

\subsection{Creating streams}\label{create_streams}
In \proglang{clRNG} library, a stream object is a structure that contains three states: the current state of the stream, the initial state (or seed) of the stream, and the initial state of the current substream. The function \fct{clrngMrg31k3pCreateStreams} from \proglang{clRNG} creates streams on the host computer using the MRG31k3p RNG. If we want to use the streams in work-items on a GPU device, the streams have to be copied from the host to the global memory of the GPU device, and then each work-item picks streams from the global memory and copies their current states only to its private memory. \pkg{gpuRandom} is able to create streams both on the host computer and on the GPU device. \fct{createStreamsCpu} is an \proglang{R} interface function of \fct{clrngMrg31k3pCreateStreams} that creates streams on CPU as \proglang{R} matrices (see the \proglang{R} output below), in which the substreams are not created because they are not actually used in random number generation. \fct{createStreamsGpu} adapted the \fct{clrngMrg31k3pCreateStreams} and creates streams directly on GPU device, so that it can avoid unnecessary data transfer between host and device. From our experience, transferring a large-size matrix of streams from host to device by converting it to a `\code{vclMatrix}' object can take very long time, hence, \fct{createStreamsGpu} is more efficient and more frequently used. In the following, we demonstrate the usage of these two functions by generating 4 streams on the host and on the device respectively.
<<createStreams_CPU>>=
# creating streams on cpu
myStreamsCpu <- createStreamsCpu(n=4, seed=12345)
t(myStreamsCpu)
@
\fct{createStreamsCpu} has two arguments:
\begin{itemize}
  \item \code{n}: number of streams to create.
  \item \code{seed}: random seed of streams of length 6, recycled if shorter. Default is set to a vector of six ``12345''.
\end{itemize}
For the MRG31k3p RNG, each state of the stream is comprised of six 31-bit integers, the state of the first stream (seed of the \proglang{clRNG} library) is by default ``12345''. We can have streams on GPU by either converting the \code{myStreamsCpu} to a `\code{vclMatrix}' or producing streams directly on GPU using \fct{createStreamsGpu}. Setting a seed same as the MRG31k3p default seed in \fct{createStreamsGpu}, we can produce on device exactly the same streams as those created on host as follows,
<<createStreams_Gpu>>=
# creating streams on gpu
myStreamsGpu1 = vclMatrix(myStreamsCpu)
myStreamsGpu2 = createStreamsGpu(n=4, seed=12345)
range(as.matrix(myStreamsGpu1 - myStreamsGpu2))
@
\fct{createStreamsGpu} also has the two arguments \code{n} and \code{seed}, \code{seed} is by default a vector of six ``12345''. \fct{createStreamsGpu} is similar to \code{.Random.seed} in \proglang{R}. By setting the RNG status with the \proglang{R} function \fct{set.seed}, we are able to repeat the sequence of random seeds, which is an important quality criteria of RNGs. Similarly in \fct{createStreamsGpu} and \fct{createStreamsCpu}, with the \code{seed} argument, we are able to produce exactly the same sequence of streams as many times as we want.

\subsection{Using streams to generate  i.i.d. $U (0, 1)$ random numbers on GPU}
The streams created on GPU are then used in work-items that executes in parallel on GPU device. In \pkg{gpuRandom} each work-item takes one distinct stream to generate random numbers, it is important to notice that the number of streams should always be equal to the number of total work-items in use. The main part of the kernel (\proglang{OpenCL} functions for execution on the device) for generating uniform random numbers is shown in Listing \ref{lst:uniformkernel}. Kernels are written in the \proglang{OpenCL C} language, in which \code{__kernel} declares a function as a kernel, pointer kernel arguments must be declared with an address space qualifier, for example \code{__global} and \code{__local}. Here the pointers to \code{streams} and to the matrix \code{out} on the global memory are passed to the kernel as arguments. \code{Nrow} and \code{Ncol} represent number of rows and columns of the matrix \code{out} respectively. \code{NpadCol} is the internal number of columns of \code{out}, \code{NpadStreams} is internal number of columns of the \code{streams} matrix (these variables are defined in macros not shown here). Each stream's current state is copied to the private memory of each work-item by the function \fct{streamsToPrivate}. \code{index} represents the position of the work-item when mapping it to a one-dimensional data structure. The function \fct{clrngMrg31k3pNextState} generates a uniform random integer named \code{temp} between 0 and 4294967295,  the value of \code{temp} is then scaled to be in the interval $(0,1)$ by multiplying it by a constant \code{mrg31k3p\_NORM\_cl}, which is defined to be $1/4294967295$ in macro not shown here. At the end of generating random numbers, streams are transferred back to global memory through the function \fct{streamsFromPrivate}.
\begin{framed}
\begin{lstlisting}[caption={captiontext},language=C,basicstyle=\small,label={lst:uniformkernel}]
 __kernel void mrg31k3pMatrix(
    __global int* streams,
    __global double* out){

@@

    }//Dcol
 }//Drow
streamsFromPrivate(streams,g1,g2,startvalue);
}
\end{lstlisting}
\end{framed}
This kernel executes over a 2-dimensional index space. Figure \ref{fig:1} illustrates how this kernel executes in practice with a simple example. Suppose we have a $2 \times 4$ NDRange ((a) in Figure \ref{fig:1}) and want to create a $4 \times 6$ matrix ((b) in Figure \ref{fig:1}) of uniform random numbers. In each grid of (a) that represents a work-item, there are two rows of numbers, the number in the above row is the stream index (from 0 to 7), and a pair of numbers in the second row is the 2-dimensional global ID of the work-item. Each stream object is passed to each work-item. The number in a cell of the matrix corresponds to the stream index, it indicates which stream does the random number generation for this matrix cell. For example, stream 0 which is allocated in work-item $(0,0)$ generates a random number in matrix $(0,0)$ in the first iteration, then moves on to matrix $(0,2)$ and generates a random number there in the second iteration, in the third iteration it generates a random number in matrix $(0,4)$ and so on until it goes out of the matrix range in the next iteration.
\begin{figure}[H]%
    \centering
    \subfloat[]{{\includegraphics[width=8cm]{/home/ruoyong/paper1_2021_5/f1} }}%
    \qquad
    \subfloat[]{{\includegraphics[width=8cm]{/home/ruoyong/paper1_2021_5/f2} }}%
    \caption{name?\label{fig:1}}%
\end{figure}

Now we use the streams created in section \ref{create_streams} to generate two vectors of 6 i.i.d. $U (0,1)$ random numbers using the function \fct{runif}. If \code{streams} and \code{Nglobal} are both missing, then by default \code{Nglobal} is set as (64,8), and a `vclMatrix' of \code{prod(Nglobal)} streams is supplied. The output is an S4 object (a `\code{vclVector}' or a `\code{vclMatrix}'), to view the generated random numbers, we need to convert it to an R vector or matrix, in this way, the random numbers are moved from the global memory of device to the host. By using \code{myStreamsGpu2} which is the same as \code{myStreamsGpu1},  \fct{runif} returns the same sequence of random numbers. 
<<simStuff2, eval=TRUE,tidy=TRUE>>=
as.vector(gpuRandom::runif(n=10, streams=myStreamsGpu1, Nglobal=c(1,nrow(myStreamsGpu1)), type = "double"))
as.vector(gpuRandom::runif(n=10, streams=myStreamsGpu2, Nglobal=c(1,nrow(myStreamsGpu1)), type = "double"))
@
The arguments of the \fct{gpuRandom::runif} are described as follows:
\begin{itemize}
  \item \code{n}: a vector of length 2 specifying the row and column number if to create a matrix, or a number specifying the length if to create a vector.
  \item \code{streams}: streams for random number generation.
  \item \code{Nglobal}: global index space. Default is set as $(64,8)$.
  \item \code{type}: ``double'' or ``float'' format of generated random numbers.
\end{itemize}
Reusing \code{myStreamsGpu1} will produce a vector different from \code{sim_1}. Because each time a random number is generated, the current state of the stream is changed (advances by one position), below shows the \code{myStreamsGpu1} after it being used.
<<eval=TRUE,tidy=TRUE>>=
t(matrix(as.matrix(myStreamsGpu1), nrow(myStreamsCpu), ncol(myStreamsCpu), dimnames = dimnames(myStreamsCpu)))
@
Unlike the objects created by \proglang{R}, by default they remain in the memory after a restart. Streams created on GPU does not remain in the memory when a new \proglang{R} session begins. However, we can save the streams on the CPU in a file using the \fct{save} function, and later recall the streams with the \fct{load} function, in this way we can reproduce the same sequence of random numbers in simulations. Below is a trivial example, in which we demonstrate saving streams to a data file called \code{streams_from_GPU.rds} on CPU and then load it back and transfer it to GPU.
<<save streams,echo=TRUE,eval=TRUE>>=
saveRDS(as.matrix(createStreamsGpu(n=4)), "myStreams.rds")
# Load the streams object as streams_saved
streams_saved <- vclMatrix(readRDS("myStreams.rds"))
@

We evaluate the distribution of 50000 GPU-generated double-precision uniform random numbers by making a density histogram  (the left panel in Figure \ref{fig0}), and plotting its empirical cumulative distribution and overlaying the theoretical cumulative distribution (the right panel in Figure \ref{fig0}). The two cumulative distribution lines seem to overlap completely, suggesting the data generated by \fct{runif} does come from a standard uniform distribution.
<<uniforms_random_vector,eval=TRUE>>=
streams <- createStreamsGpu(n = 512)
random_u<-runif(n=50000, streams=streams, Nglobal=c(64,8), type="double")
random_u<-as.vector(random_u)
@
\begin{figure}[H]
\centering
<<largenumberuniforms, eval=TRUE, echo=FALSE, fig.height=4, fig.width=4, fig.show='hold'>>=
hist(random_u, prob = TRUE, main="Histogram")
plot(ecdf(random_u), ylab="F(x)",main="Empirical cumulative distribution")
curve(punif(x, 0, 1), lty=2, add=TRUE, col="red")
@
\caption{Histogram and empirical cumulative distribution of the GPU-generated random numbers.\label{fig0}}
\end{figure}

\subsection{Using streams to generate i.i.d. N (0, 1) random numbers on GPU}
We apply the Box-Muller transformation to $U (0,1)$ random numbers to generate Gaussian random numbers. As shown in Algorithm \ref{algorithm1}, Box-Muller algorithm takes two independent, standard uniform random variables $U_1$ and $U_2$ and produces two independent, standard Gaussian random variables $X$ and $Y$, where $R$ and $\Theta$ are polar coordinate random variables. The derivation basically uses transformation from Cartesian coordinates to polar coordinates to represent two independent, Normally distributed random variables. The Box-Muller algorithm turns out to be the best choice for Gaussian transform on GPU compared to other transform methods \citep{howes2007efficient}, because this algorithm has no branching or looping, which are the things GPU is not good at, but GPU is good at the high computational load of sine and cosine functions in the algorithm.

@@

Listing \ref{lst:normal} shows a fragment of the kernel that generates standard Gaussian random numbers. Aside from using local memory and the part that does the Box-Muller transformation, this kernel is the same with the kernel for generating uniform random numbers shown in Listing \ref{lst:uniformkernel}. If developers want to use other Gaussian transform methods, this kernel is easy to be modified and incorporated in other \proglang{R} packages. The kernel executes over a 2-dimensional index space of $1 \times 2$ work-groups. \code{part[0]} and \code{part[1]} correspond to $R$ and $\Theta$ in the formulas respectively. As the work-items in a work-group proceed at differing rates, \code{barrier(CLK_LOCAL_MEM_FENCE)} ensures correct ordering of memory operations to local memory, so that Gaussian random numbers $(X_1,Y_1), \dots, (X_n,Y_n)$ are generated in pairs correctly, errors such as $(X_n, Y_{(n-1)})$ or $(X_{(n-1)}, Y_{(n)})$ are avoided.
%
\begin{framed}
\begin{lstlisting}[language=C,basicstyle=\small,caption={name},label={lst:normal}]
const int 
index = get_global_id(0)*get_global_size(1) + get_global_id(1);
const int 
DrowInc = 2*get_global_size(0), DrowStartInc = DrowInc * NpadCol;
int Drow, Dcol, DrowStart, Dentry;
uint temp;
uint g1[3], g2[3];
const int startvalue=index * NpadStreams;
local double part[2], cosPart1, sinPart1;

@@

Figure \ref{fig4} visualizes the approximate sampling distributions of the test statistics from the two examples "Month" and "Week", by plotting a histogram for each of the example. The values of observed statistics are calculated by \fct{gpuRandom::logfactSum} and are marked with a blue line on each plot. 
\begin{figure}[H]
\centering
<<fighistMonth, eval=TRUE, echo=FALSE, fig.show='hold', cache=TRUE>>=
hist(month_stats, xlab=TeX('-$\\sum(\\log(n_{ij}))$'), breaks=40, ylab="Frequency", main="Month data")
abline(v = observed_m, col = "blue", lwd = 1.5)

>>>>

\title{Random number generation and applications on GPU’s in R(?)}
\Plaintitle{}
\Shorttitle{}

@@

In recent years, parallel computing with \proglang{R} \citep{r2021} has become a very important topic and attracted lots of interest from researchers \citep[see][for a review]{eddelbuettel2021parallel}. Although \proglang{R} is one of the most popular statistical software with many advantages, it has drawbacks in memory usage and computation mode aspects \citep{zhao_2016}. To be more specific, (1) \proglang{R} requires all data to be loaded into the main memory (RAM) and thus can handle a very limited size of data; (2) \proglang{R} is a single-threaded program, it can not effectively use all the computing cores of multi-core processors. Parallel computing is the solution to these drawbacks of \proglang{R}, for an overview of current parallel computing approaches with \proglang{R}, see CRAN Task View by \citet{cran2021} at \url{https://cran.r-project.org/web/views/HighPerformanceComputing.html}.

An important way of parallel computing with \proglang{R} is using Graphics Processing Units (GPUs). GPUs can perform thousands of tests simultaneously, which makes them powerful at doing massive parallel computing and they are relatively cheap compared to multicore CPU's. Although there has been a few \proglang{R} packages developed that provide some GPU capability in the past years, they come with some restrictions. Packages such as \pkg{gputools}, \pkg{gpumatrix}, \pkg{cudaBayesreg}, \pkg{rpud} (available on github), are now no longer maintained, the popular \pkg{tensorflow} \citep{tensorflow1} package uses GPU via \proglang{Python}, which makes it hard to be incorporated in a new \proglang{R} package that uses GPU. All of these mentioned packages are restricted to \proglang{R} users with NVIDIA GPUs. \pkg{gpuR} \citep{gpur1} is the only \proglang{R} package with a convenient interface between \proglang{R} and GPUs and it is created for any \proglang{R} user with a GPU device. By utilizing the \pkg{ViennaCL} \citep*{rupp2016viennacl} library, it provides a bridge between \proglang{R} and non-proprietary GPUs through the \proglang{OpenCL} (Open Computing Language) backend, which when combined with \pkg{Rcpp} \citep{rcpp1} gives a building block for other \proglang{R} packages.

Random number generation is critical in simulation-based statistical inference, machine learning and many other scientific fields. While most random number generators are sequential, the \proglang{R} packages \pkg{parallel} (not on cran, how to cite?) and \pkg{future} \citep*{future1.19.1} are able to generate random numbers in parallel on multicore CPUs. More specifically, \pkg{parallel} writes an interface for the \pkg{RngStreams}, a \proglang{\texttt{C++}} library by \cite{l2002object} which is based on a combined multiple-recursive generator (MRG), \pkg{future} also uses the combined MRG algorithm for generating random numbers, however, it is not very clear how the streams are implemented (?).

There is not yet an \proglang{R} package that is able to generate random numbers on GPU, this is complicated to do because multiple streams will be needed for parallel generation of random numbers, and the stream states need to be able to be restored for reproducibility of results, which is essential in many simulation applications. We introduce the \proglang{R} package \pkg{gpuRandom} that leverages the \pkg{gpuR} package, and can generate random numbers on GPU by using the \pkg{clRNG} \citep{l2015clrng} library, and can thus accelerate several types of statistical simulation and modelling.

The remaining sections are organized as follows:
In Section 2, we introduce streams and the use of streams in work-items on GPU device, and describe and check the functions for generating uniform and normal random numbers.
In Section 3, we apply GPU-generated uniform random numbers in Monte Carlo simulation for Fisher’s exact test, we parallelise this Monte Carlo simulation and implement it on GPU, then we provide two real data examples to demonstrate the function usage and test its R performance.
Section 4 simulates simultaneously multiple Gaussian random surfaces by applying GPU-generated Normal random numbers and GPU-computed batches of Mat\'ern covariance matrices, then we plot the generated random surfaces in the end. 
Finally, the paper concludes with a short summary and a discussion in Section 5.

@@

\section{Generating random numbers} \label{}
Uniform random number generators (RNGs) are essential in simulating random numbers from all types of probability distributions. \cite{l2012random} summarized the usual two steps to generate a random variable in computational statistics: (1) generating independent and identically distributed (i.i.d.) uniform random variables on the interval $(0, 1)$, (2) applying transformations to these i.i.d. $U(0, 1)$ random variables to get samples from the desired distribution. \cite{l2012random} and \cite{robert2004random} present many general transformation methods for generating non-uniform random variables, for example, the most frequently used inverse transform method, the Box-Muller algorithm \citep{box1958note} for Gaussian random variable generation, and so on, which are all built on uniform random variables. %In the rest paper, ``RNG'' refers to uniform pseudo random number generator.

\pkg{clRNG} is an \proglang{OpenCL} library for uniform random number generation, it provides four different RNGs: the MRG31k3p, MRG32k3a, LFSR113, and Philox-4×32-10 generators. \pkg{gpuRandom} package uses the MRG31k3p RNG from the \proglang{clRNG}, making it able to generate random numbers on GPUs. In what follows, we will illustrate how to create streams and how to use streams to generate (standard) uniform and (standard) Gaussian random numbers.

\subsection{Creating streams}\label{create_streams}
The popular way of obtaining multiple streams is to take an RNG with a long period and cut the RNG's sequence into very long pieces, these pieces are adjacent and of equal length $Z$, so each piece can be used as a separate stream. Creating a new stream amounts to computing its starting point (seed). Each of these streams can also be partitioned into substreams with equally-spaced starting points \citep{l2002object}, which are occasionally useful.

In general, a stream object contains three states: the current state of the stream， the initial state of the stream (or seed), and the initial state of the current substream (by default it is equal to the seed). Whenever the user creates a new stream, the software automatically jumps ahead by Z steps to find its initial state, and the three states in the stream object are set to it.

In \proglang{clRNG} library, the initial state of the first stream has a default value (12345). For the MRG31k3p RNG, the entire period length of approximately $2^{185}$ is divided into approximately $2^{51}$ non-overlapping streams of length $Z = 2^{134}$. Each stream is further partitioned into substreams of length $W = 2^{72}$. The state (and seed) of each stream is a vector of six 31-bit integers. This size of state is appropriate for having streams running in work items on GPU cards, for example, while providing a sufficient period length for most applications.

The function \fct{clrngMrg31k3pCreateStreams} from \proglang{clRNG} creates streams on the host using the MRG31k3p RNG. If we want to use the streams in work-items on a GPU device, the streams have to be copied from the host to the global memory of the GPU device, and then each work-item copies the current states only of the streams to its private memory. \pkg{gpuRandom} is able to create streams both on the host and on the GPU device. \fct{createStreamsCpu} is an interface function of \fct{clrngMrg31k3pCreateStreams} that creates streams on host as \proglang{R} matrices (see the \proglang{R} output below), creating streams on host makes it easy for us to view and check the streams behavior.  \fct{createStreamsGpu} is adapted from the \fct{clrngMrg31k3pCreateStreams} and creates streams directly on a GPU device. Substreams are not created because currently we use one stream per work-item. %(From our experience, transferring a large-size matrix of streams from host to device by converting it to a `\code{vclMatrix}' object can take very long time, hence, \fct{createStreamsGpu} is more efficient and more frequently used. delete?) 
In the following, we demonstrate the usage of these two functions by generating 4 streams on the host and on the device respectively. 
<<createStreams_CPU>>=
# creating streams on CPU
library("gpuR")
library("gpuRandom")
myStreamsCpu <- createStreamsCpu(n=4, seed=12345)
t(myStreamsCpu)
@
\fct{createStreamsCpu} has two arguments:
\begin{itemize}
  \item \code{n}: number of streams to create.
  \item \code{seed}: vector of length 6, recycled if shorter. Default is set to a vector of six ``12345''.
\end{itemize}
We can have streams on GPU by either converting the \code{myStreamsCpu} to a `\code{vclMatrix}' or producing streams directly on GPU using \fct{createStreamsGpu}. Setting a seed same as the MRG31k3p default seed in \fct{createStreamsGpu}, we can produce on device exactly the same streams as those created on host as follows,
<<createStreams_Gpu>>=
# creating streams on GPU
myStreamsGpu1 = vclMatrix(myStreamsCpu)
myStreamsGpu2 = createStreamsGpu(n=4, seed=12345)
range(as.matrix(myStreamsGpu1 - myStreamsGpu2))
@
\fct{createStreamsGpu} also has the two arguments \code{n} and \code{seed}, \code{seed} is by default a vector of six ``12345''. \fct{createStreamsGpu} is similar to \code{.Random.seed} in \proglang{R}. By setting the RNG status with the \proglang{R} function \fct{set.seed}, we are able to repeat the sequence of random seeds, which is an important quality criteria of RNGs. Similarly in \fct{createStreamsGpu} and \fct{createStreamsCpu}, with the \code{seed} argument, we make the sequence of streams repeatable.

\subsection{Using streams to generate  i.i.d. $U (0, 1)$ random numbers on GPU}
The streams created sequentially on a GPU are then used in work-items that executes in parallel on a GPU device. In \pkg{gpuRandom} each work-item takes one distinct stream to generate random numbers, so the number of streams should always be equal to the number of total work-items in use. The main part of the kernel (\proglang{OpenCL} functions for execution on the device) for generating uniform random numbers is shown in Listing \ref{lst:uniformkernel}. Kernels are written in the \proglang{OpenCL C} language, in which \code{__kernel} declares a function as a kernel, pointer kernel arguments must be declared with an address space qualifier, for example \code{__global} and \code{__local}. Here the pointers to \code{streams} and to the output matrix \code{out} on the global memory are passed to the kernel as arguments. \code{Nrow} and \code{Ncol} represent row and column number of the matrix \code{out} respectively. \code{NpadCol} is the internal number of columns of \code{out}, \code{NpadStreams} is the internal number of columns of the \code{streams} (these variables are defined in macros not shown here). Each stream's current state is copied to the private memory of each work-item by the function \fct{streamsToPrivate}. \code{index} represents the position of the work-item when mapping it to a one-dimensional data structure. The function \fct{clrngMrg31k3pNextState} generates an uniform random integer called \code{temp} between 0 and 4294967295,  the value of \code{temp} is then scaled to be in the interval $(0,1)$ by multiplying it by a constant \code{mrg31k3p\_NORM\_cl}, which is defined to be $1/4294967295$ in macro not shown here. At the end of generating random numbers, streams are transferred back to global memory through the function \fct{streamsFromPrivate}.
\begin{framed}
\begin{lstlisting}[caption={captiontext},language=C,basicstyle=\small,label={lst:uniformkernel}]
 __kernel void mrg31k3pMatrix(
    __global int* streams,
    __global double* out){

@@

    }//Dcol
 }//Drow
streamsFromPrivate(streams,g1,g2,startvalue);
}
\end{lstlisting}
\end{framed}
This kernel executes over a 2-dimensional NDRange. Figure \ref{fig:1} illustrates how this kernel executes in practice with a simple example. Suppose we have a $2 \times 4$ NDRange ((a) in Figure \ref{fig:1}) and want to create a $4 \times 6$ matrix ((b) in Figure \ref{fig:1}) of uniform random numbers. In each grid of (a) that represents a work-item, there are two rows of numbers, the number in the above row is the stream index (from 0 to 7), and a pair of numbers in the second row is the 2-dimensional global ID of the work-item. Each stream object is passed to each work-item. The number in a cell of the matrix corresponds to the stream index, it indicates which stream generates a random number for this matrix cell. For example, stream 0 which is allocated in work-item $(0,0)$ generates a random number in matrix $(0,0)$ in the first iteration, then moves on to matrix $(0,2)$ and generates a random number there in the second iteration, in the third iteration it generates a random number in matrix $(0,4)$ and so on until it goes out of the matrix range in the next iteration.
\begin{figure}[H]%
    \centering
    \subfloat[]{{\includegraphics[width=8cm]{/home/ruoyong/diseasemapping/pkg/gpuRandom/inst/documents/paper1_2021_5/f1} }}%
    \qquad
    \subfloat[]{{\includegraphics[width=8cm]{/home/ruoyong/diseasemapping/pkg/gpuRandom/inst/documents/paper1_2021_5/f2} }}%
    \caption{name?\label{fig:1}}%
\end{figure}

Now we use the streams created in section \ref{create_streams} to generate two vectors of double-precison i.i.d. $U(0,1)$ random numbers with the function \fct{runif}. To view the generated random numbers, we need to convert them to R vectors or matrices, by doing this, the random numbers are moved from the global memory of the GPU device to the host.
<<simStuff2, eval=TRUE,tidy=TRUE>>=
as.vector(gpuRandom::runif(n=6, streams=myStreamsGpu1, Nglobal=c(1,nrow(myStreamsGpu1)), type = "double"))

as.vector(gpuRandom::runif(n=10, streams=myStreamsGpu2, Nglobal=c(1,nrow(myStreamsGpu1)), type = "double"))[-(1:6)]
@
The arguments of the \fct{gpuRandom::runif} are described as follows:
\begin{itemize}
  \item \code{n}: a vector of length 2 specifying the row and column number if to create a matrix, or a number specifying the length if to create a vector.
  \item \code{streams}: streams for random number generation.
  \item \code{Nglobal}: global index space. Default is set as $(64,8)$.
  \item \code{type}: ``double'' or ``float'' or ``integer'' format of generated random numbers.
\end{itemize}
\fct{runif} returns the same sequence of random numbers if it uses the same streams in the same current state (here \code{myStreamsGpu1} and \code{myStreamsGpu2}), not streams starting from the same states. Reusing \code{myStreamsGpu1} will produce a vector different from \code{sim_1}. Because each time a random number is generated, the current state of the stream advances by one position. Below shows the \code{myStreamsGpu1} after it being used.
<<eval=TRUE,tidy=TRUE, include=TRUE>>=
t(matrix(as.matrix(myStreamsGpu1), nrow(myStreamsCpu), ncol(myStreamsCpu), dimnames = dimnames(myStreamsCpu)))
@
Unlike the objects created by \proglang{R}, by default they remain in the memory after a restart. Streams created on GPU does not remain in the memory when a new \proglang{R} session begins. However, we can save the streams on the CPU in a file using the \fct{saveRDS} function, and later recall the streams with the \fct{readRDS} function, in this way we can reproduce the exact same sequence of random numbers in simulations. Below is a trivial example, in which we demonstrate saving streams to a data file called \code{streams_from_GPU.rds} on CPU and then load it back and transfer it to GPU.
<<save streams,echo=TRUE,eval=TRUE>>=
saveRDS(as.matrix(createStreamsGpu(n=4)), "myStreams.rds")
# Load the streams object as streams_saved
streams_saved <- vclMatrix(readRDS("myStreams.rds"))
@

We evaluate the distribution of 50000 GPU-generated double-precision uniform random numbers by making a density histogram  (the left panel in Figure \ref{fig0}), and plotting its empirical cumulative distribution and overlaying the theoretical cumulative distribution (the right panel in Figure \ref{fig0}). The two cumulative distribution lines seem to overlap completely, suggesting the data generated by \fct{runif} does come from a standard uniform distribution.
<<uniforms_random_vector,eval=TRUE>>=
streams <- createStreamsGpu(n = 512)

random_u<-as.vector(runif(n=50000, streams=streams, Nglobal=c(64,8), type="double"))
@
\begin{figure}[H]
\centering
<<largenumberuniforms, eval=TRUE, echo=FALSE, fig.height=4, fig.width=4, fig.show='hold'>>=
hist(random_u, prob = TRUE, main="Histogram")
plot(ecdf(random_u), ylab="F(x)",main="Empirical cumulative distribution")
curve(punif(x, 0, 1), lty=2, add=TRUE, col="red")
@
\caption{Histogram and empirical cumulative distribution of the GPU-generated random numbers.\label{fig0}}
\end{figure}

\subsection{Using streams to generate i.i.d. N(0, 1) random numbers on GPU}
We apply the Box-Muller transformation to $U(0,1)$ random numbers to generate standard Gaussian random numbers. As shown in Algorithm \ref{algorithm1}, Box-Muller algorithm takes two independent, standard uniform random variables $U_1$ and $U_2$ and produces two independent, standard Gaussian random variables $X$ and $Y$, where $R$ and $\Theta$ are polar coordinate random variables. %The derivation basically uses transformation from Cartesian coordinates to polar coordinates to represent two independent, Normally distributed random variables. 
The Box-Muller algorithm turns out to be the best choice for Gaussian transform on GPU compared to other transform methods \citep{howes2007efficient}, because this algorithm has no branching or looping, which are the things GPU is not good at. %but GPU is good at the high computational load of sine and cosine functions in the algorithm.

@@

Listing \ref{lst:normal} shows a fragment of the kernel that generates standard Gaussian random numbers. Aside from using local memory and the part that does the transformation, this kernel is the same with the kernel for generating uniform random numbers shown in Listing \ref{lst:uniformkernel}. If developers want to use other Gaussian transform methods, this kernel is easy to be adapted and incorporated in other \proglang{R} packages. The kernel executes over a 2-dimensional NDRange with $1 \times 2$ work-groups. \code{part[0]} and \code{part[1]} correspond to $R$ and $\Theta$ in the formulas respectively. As the work-items in a work-group proceed at differing rates, \code{barrier(CLK_LOCAL_MEM_FENCE)} ensures correct ordering of memory operations to local memory, so that Gaussian random numbers $(X_1,Y_1), \dots, (X_n,Y_n)$ are generated in pairs correctly, errors such as $(X_n, Y_{(n-1)})$ or $(X_{(n-1)}, Y_{(n)})$ are avoided.
%
\begin{framed}
\begin{lstlisting}[language=C,basicstyle=\small,caption={name},label={lst:normal}]
const int 
index = get_global_id(0)*get_global_size(1) + get_global_id(1);
const int 
DrowInc = 2*get_global_size(0), DrowStartInc = DrowInc * NpadCol;
int Drow, Dcol, DrowStart, Dentry;
uint temp;
uint g1[3], g2[3];
const int startvalue=index * NpadStreams;
local double part[2], cosPart1, sinPart1;

@@

Figure \ref{fig4} visualizes the approximate sampling distributions of the test statistics from the two examples "Month" and "Week", by plotting a histogram for each of the example. The values of observed statistics are calculated by \fct{gpuRandom::logfactSum} and are marked with a blue line on each plot. 
\begin{figure}[H]
\centering
<<fighistMonth, eval=TRUE, echo=FALSE, fig.show='hold', cache=TRUE>>=
hist(month_stats, xlab=TeX('-$\\sum(\\log(n_{ij}!))$'), breaks=40, ylab="Frequency", main="Month data")
abline(v = observed_m, col = "blue", lwd = 1.5)

hist(result_weekgpu$cpu, xlab=TeX('-$\\sum(\\log(n_{ij}!))$'), ylab="Frequency", main="Week data")
abline(v = observed_w, col = "blue", lwd = 1.5)
@
\caption{ Frequency distribution for test statistics from month and week data tables.\label{fig4}}
\end{figure}
